Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-01 paper/paper.qmd
Write performance results section

---
title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
bibliography: references.bib
fontsize: 11pt
date: last-modified
date-format: medium
prefer-html: true
format:
  manuscript-pdf:
    mainfont: Helvetica
    keep-tex: true
    linenumbers: true
    doublespacing: true
    lof: false
  html:
    embed-resources: true
runningtitle: Predicting CDI Severity from OTUs
runningauthor: Sovacool
papertype: Preprint
author:
  - name: Kelly L. Sovacool
    orcid: 0000-0003-3283-829X
    affiliations:
      - ref: bioinf
  - name: Sarah E. Tomkovich
    affiliations:
      - ref: micro
  - name: Megan L. Coden
    affiliations:
      - ref: mcdb
  - name: Vincent B. Young
    affiliations:
      - ref: micro
      - ref: intmed
  - name: Krishna Rao
    affiliations:
      - ref: intmed
  - name: Patrick D. Schloss
    orcid: 0000-0002-6935-4275
    email: pschloss@umich.edu
    corresponding: true
    affiliations:
      - ref: micro
      - ref: ccmb
affiliations:
  - id: bioinf
    name: Department of Computational Medicine & Bioinformatics, University of Michigan
  - id: micro
    name: Department of Microbiology & Immunology, University of Michigan
  - id: mcdb
    name: Department of Molecular, Cellular, and Developmental Biology, University of Michigan
  - id: intmed
    name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
  - id: ccmb
    name: Center for Computational Medicine and Bioinformatics, University of Michigan
#abstract: TODO
#importance: TODO
keywords: _C. difficile_ infection, supervised machine learning, gut microbiome, amplicion sequencing
---
```{r setup, eval=TRUE, echo=FALSE, cache=FALSE}
schtools::set_knitr_opts()
knitr::opts_chunk$set(dpi = 600)
```
```{r load_data, message = FALSE}
library(here)
library(kableExtra)
library(knitr)
library(tidyverse)

@@

We first set out to train the best models possible for each severity definition.
Not all samples have labels available for all four severity definitions due to
missing data for some patient lab values and incomplete chart review (@fig-flowchart B),
thus each severity definition has a different number of samples (@tbl-counts)
when using as many samples as possible. We refer to these as the full datasets.
Random forest models were trained on repeated 100 splits of the datasets into
training and test sets, and performance was evaluated on the held-out
test set using the area under the receiver-operator characteristic curve (AUROC)
the area under the balanced precision-recall curve (AUBPRC).

However, comparisons across these definitions is not fair when using different
subsets of the data for each definition.
To better compare the model performances across different severity definitions,
we selected the intersection of samples (n=`r n_cases_int`) that had labels for
all four severity definitions and repeated the model training and evaluation
process.

Report median AUROC for training set and test set, and median AUBPRC for test set (@fig-performance).
Nearly all pairs of definitions have significantly different performances on the
test set (P < 0.05) except for AUROC and AUBPRC of Attributable vs. Pragmatic on
the intersection dataset (as they are identical), AUROC of Attributable vs.
All-cause on the full dataset, and AUROC of Attributable vs. IDSA on the full
dataset.

@@

To calculate a clinically-relevant NNS for these models, we computed the
confusion matrix at the 95th percentile of risk for each prediction model
(@tbl-risk).
Among the models predicting severe outcomes, those trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition.
For context, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data and from a smaller set of clinician-curated
variables, achieving precision values of 0.417 (NNS = `r ehr_nns`) for the EHR
model and 0.167 (NNS = `r curated_nns`) for the curated model at the 95th
percentile of risk [@li_using_2019;@rao_clostridium_2015].
Multiplying the NNS of the OTU-based models by the estimated NNT of 
`r fdx_nnt` for fidaxomicin yields NNB values of `r allcause_fdx_nnb` 
for all-cause severity, `r attrib_fdx_nnb` for attributable severity,
and `r pragmatic_fdx_nnb` for pragmatic severity.
Thus, in a hypothetical scenario where these assumptions about fidaxomicin were
correct, between `r min_fdx_nnb` and `r max_fdx_nnb` patients would need to be
predicted to experience a severe outcome and be treated with fidaxomicin in
order for one patient to benefit.
As the NNS values were computed at the 95th percentile of risk (where 5% of
patients screened are predicted to experience severity), these NNB values mean
that `r min_screen` to `r max_screen` total CDI patients would need to be
screened by an OTU-based prediction model in order for one patient to benefit.
For comparison, pairing the prior EHR-based model with fidaxomicin would yield
an NNB of `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened for
one patient to benefit.
These estimates represent a proof-of-concept demonstration of the potential
value of deploying severity prediction models to guide clinicians' treatment
decisions.

@@

::: {#fig-performance fig-cap="**Performance of ML models.**
In the left panels, models were trained on the full dataset, with different
numbers of samples available for each severity definition.
In the right panels, models were trained on the intersection of samples with all
labels available for each definition.
Note that the intersection dataset for Attributable and Pragmatic severity have
exactly the same labels, thus identical values are expected.
**A)** 
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets, and the area under the balanced
precision-recall curve (AUBPRC) for the test sets.
Each point is the median performance across 100 train/test splits with tails as
the 95% CI.
Nearly all pairs of definitions have significantly different performances on the
test set (P < 0.05) except for AUROC and AUBPRC of Attributable vs. Pragmatic on
the intersection dataset (as they are identical), AUROC of Attributable vs.
All-cause on the full dataset, and AUROC of Attributable vs. IDSA on the full
dataset.
**B)** Receiver-operator characteristic curves for the test sets.
Mean specificity is reported at each sensitivity value, 
with ribbons as the 95% CI.
**C)** Balanced precision-recall curves for the test sets.
Mean balanced precision is reported at each recall value, 
with ribbons as the 95% CI.
"}
![](figures/ml-performance.png)
:::

>>>>

---
title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
bibliography: references.bib
fontsize: 11pt
date: last-modified
date-format: medium
prefer-html: true
format:
  manuscript-pdf:
    mainfont: Helvetica
    keep-tex: true
    linenumbers: true
    doublespacing: true
    lof: false
  html:
    embed-resources: true
runningtitle: Predicting CDI Severity from OTUs
runningauthor: Sovacool
papertype: Preprint
author:
  - name: Kelly L. Sovacool
    orcid: 0000-0003-3283-829X
    affiliations:
      - ref: bioinf
  - name: Sarah E. Tomkovich
    affiliations:
      - ref: micro
  - name: Megan L. Coden
    affiliations:
      - ref: mcdb
  - name: Vincent B. Young
    affiliations:
      - ref: micro
      - ref: intmed
  - name: Krishna Rao
    affiliations:
      - ref: intmed
  - name: Patrick D. Schloss
    orcid: 0000-0002-6935-4275
    email: pschloss@umich.edu
    corresponding: true
    affiliations:
      - ref: micro
      - ref: ccmb
affiliations:
  - id: bioinf
    name: Department of Computational Medicine & Bioinformatics, University of Michigan
  - id: micro
    name: Department of Microbiology & Immunology, University of Michigan
  - id: mcdb
    name: Department of Molecular, Cellular, and Developmental Biology, University of Michigan
  - id: intmed
    name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
  - id: ccmb
    name: Center for Computational Medicine and Bioinformatics, University of Michigan
#abstract: TODO
#importance: TODO
keywords: _C. difficile_ infection, supervised machine learning, gut microbiome, amplicion sequencing
---
```{r setup, eval=TRUE, echo=FALSE, cache=FALSE}
schtools::set_knitr_opts()
knitr::opts_chunk$set(dpi = 600)
```
```{r load_data, message = FALSE}
library(assertthat)
library(here)
library(kableExtra)
library(knitr)
library(tidyverse)

@@

We first set out to train the best models possible for each severity definition.
Not all samples have labels available for all four severity definitions due to
missing data for some patient lab values and incomplete chart review
(@fig-flowchart B), thus each severity definition has a different number of
samples when using as many samples as possible (@tbl-counts).
We refer to these as the full datasets.
Random forest models were trained on 100 splits of the datasets into training
and test sets, and performance was evaluated on the held-out test set using the
area under the receiver-operator characteristic curve (AUROC).
Since the severity classes are highly imbalanced with different proportions of
severe samples between definitions, we also calculated the balanced precision
and the area under the balanced precision-recall curve (AUBPRC) as first
proposed by @wu_improved_2021 to describe the precision that would be expected
if the outcome classes were balanced.

```{r model_comps_full, output = FALSE}
model_comps <- read_csv(here('results','model_comparisons.csv'))
comps_full_auroc <- model_comps %>% 
  filter(dataset == 'full', metric == 'AUROC', p_value < 0.05)
assert_that(all((comps_full_auroc %>% pull(group1)) == c("allcause", "attrib", "idsa", "idsa")))
assert_that(all((comps_full_auroc %>% pull(group2)) == c("pragmatic", "pragmatic", "allcause", "pragmatic")))
assert_that(all(
  model_comps %>%
    filter(dataset == 'full', metric == 'AUBPRC', p_value > 0.05) %>%
    select(group1, group2) %>%
    as.vector() %>% 
    unlist() == c('attrib', 'allcause')
))
```

After training on the full datasets, the performance as measured by the AUROCs
of the training set cross-validation folds were similar to those of the held-out
test sets, indicating that the models are neither overfit nor underfit
(@fig-performance A).
As measured by AUROC on the held-out test sets, models predicting pragmatic
severity performed best with a median AUROC of `r perf_medians$auroc_full_pragmatic`,
and this was significantly different from that of the other definitions 
on the full datasets (P < 0.05).
Models predicting IDSA, all-cause, and attributable severity performed similarly
with median test set AUROCs of
`r perf_medians$auroc_full_idsa`,
`r perf_medians$auroc_full_allcause`, and
`r perf_medians$auroc_full_attrib` respectively.
The test set AUROCs were not significantly different (P > 0.05) for attributable
and IDSA nor for attributable and all-cause, but the IDSA and all-cause AUROCs
were significantly different from each other (P < 0.05).
We plotted the receiver-operator characteristic curve and found that the
pragmatic severity models outperformed the others at all specificity values
(@fig-performance B).
A prior study trained a logistic regression model on whole Electronic Health
Record data extracted on the day of CDI diagnosis to predict attributable
severity, yielding an AUROC of `r ehr_auroc` [@li_using_2019].
While our OTU-based attributable severity model did not meet this performance,
the OTU-based pragmatic severity model performed just as well as the EHR-based
model in terms of AUROC.

The test set median AUBPRCs from the full datasets followed a similar pattern as 
the test set AUROCs with
`r perf_medians$aubprc_full_idsa` for IDSA severity,
`r perf_medians$aubprc_full_allcause` for all-cause severity,
`r perf_medians$aubprc_full_attrib` for attributable severity, and
`r perf_medians$aubprc_full_pragmatic` for pragmatic severity.
The AUBPRCs were significantly different from each other (P < 0.05) for
each pair of severity definitions except for attributable vs all-cause.
We plotted the balanced precision-recall curve and found that the IDSA definition
outperformed all other models at very low recall values, but the others outperform
IDSA at all other points of the curve (@fig-performance C).
The 95% confidence intervals overlapped the baseline AUROC and AUBPRC for
the attributable severity models, while all others did not overlap the baseline.

While it is advantageous to use as much data as available to train the best
models possible, comparing performances of models trained on different subsets
of the data is not entirely fair.
To enable fair comparisons of the model performances across different severity
definitions, we also selected the intersection of samples (n=`r n_cases_int`)
that had labels for all four severity definitions and repeated the model
training and evaluation process on this intersection dataset.
The attributable definition is exactly the same as the pragmatic definition for
the intersection dataset, as we defined pragmatic severity to use the
attributable label when available.
The performance results on the intersection dataset are shown in the right
facets of each panel of @fig-performance.

```{r model_comps_int, output = FALSE}

comps_int_auroc <- model_comps %>% filter(group1 != 'pragmatic', group2 != 'pragmatic') %>% 
  filter(dataset == 'int', metric == 'AUROC', p_value < 0.05)
assert_that(all(comps_int_auroc %>% pull(group1) == c('attrib', 'idsa')))
assert_that(all(comps_int_auroc %>% pull(group2) == 'allcause'))

comps_int_aubprc <- model_comps %>% filter(group1 != 'pragmatic', group2 != 'pragmatic') %>% 
  filter(dataset == 'int', metric == 'AUBPRC', p_value < 0.05)

assert_that(all(comps_int_aubprc %>% pull(group1) == c('attrib', 'idsa')))
assert_that(all(comps_int_aubprc %>% pull(group2) == 'allcause'))
```

As with the full datasets, the AUROCs of the training sets and test sets were
similar within each severity definition.
The median test set AUROCs were 
`r perf_medians$auroc_int_idsa` for IDSA severity,
`r perf_medians$auroc_int_allcause` for all-cause severity,
`r perf_medians$auroc_int_attrib` and for attributable severity.
The AUROCs on the intersection dataset were significantly different for
all-cause vs attributable and all-cause vs IDSA severity (P < 0.05), but not for
IDSA vs attributable severity (P > 0.05).
The median test set AUBPRCs were
`r perf_medians$aubprc_int_idsa` for IDSA severity,
`r perf_medians$aubprc_int_allcause` for all-cause severity,
`r perf_medians$aubprc_int_attrib` and for attributable severity.
Just as with the AUROCs, the AUBPRCs were significantly different for
all-cause vs attributable and all-cause vs IDSA severity (P < 0.05), but not for
IDSA vs attributable severity (P > 0.05).
For all severity definitions, performance dropped between the full dataset and
the intersection dataset since fewer samples are available, but this effect is
least dramatic for IDSA severity as the full and intersection datasets are more
similar for this definition.
The 95% confidence interval overlaps with the baseline for both AUROC and AUBPRC
for all definitions on the intersection dataset except for IDSA severity.

@@

To calculate a clinically-relevant NNS for these models, we computed the
confusion matrix at the 95th percentile of risk for each prediction model
(@tbl-risk).
Among the models predicting severe outcomes, those trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition.
For context, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data extracted two days after diagnosis and from a
smaller set of manually curated variables, achieving precision values of 0.417
(NNS = `r ehr_nns`) for the EHR model and 0.167 (NNS = `r curated_nns`) for the
curated model at the 95th percentile of risk
[@li_using_2019;@rao_clostridium_2015]. <!-- TODO what about NNS on day of diagnosis? -->
Multiplying the NNS of the OTU-based models by the estimated NNT of 
`r fdx_nnt` for fidaxomicin yields NNB values of `r allcause_fdx_nnb` 
for all-cause severity, `r attrib_fdx_nnb` for attributable severity,
and `r pragmatic_fdx_nnb` for pragmatic severity.
Thus, in a hypothetical scenario where these assumptions about fidaxomicin hold
true, between `r min_fdx_nnb` and `r max_fdx_nnb` patients would need to be
predicted to experience a severe outcome and be treated with fidaxomicin in
order for one patient to benefit.
As the NNS values were computed at the 95th percentile of risk (where 5% of
patients screened are predicted to experience severity), these NNB values mean
that `r min_screen` to `r max_screen` total CDI patients would need to be
screened by an OTU-based prediction model in order for one patient to benefit.
For comparison, pairing the prior EHR-based model with fidaxomicin would yield
an NNB of `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened for
one patient to benefit.
These estimates represent a proof-of-concept demonstration of the potential
value of deploying severity prediction models to guide clinicians' treatment
decisions.

@@

::: {#fig-performance fig-cap="**Performance of ML models.**
In the left facets, models were trained on the full datasets, with different
numbers of samples available for each severity definition.
In the right facets, models were trained on the same dataset consisting of the
intersection of samples with all labels available for each definition.
Note that the intersection dataset for Attributable and Pragmatic severity have
exactly the same labels, thus they have identical performance.
**A)** 
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets, and the area under the balanced
precision-recall curve (AUBPRC) for the test sets.
Each point is the median performance across 100 train/test splits with tails as
the 95% CI.
Nearly all pairs of definitions have significantly different performances on the
test set (P < 0.05) except for AUROC and AUBPRC of Attributable vs. Pragmatic on
the intersection dataset (as they are identical), AUROC of Attributable vs.
All-cause on the full dataset, and AUROC of Attributable vs. IDSA on the full
dataset.
**B)** Receiver-operator characteristic curves for the test sets.
Mean specificity is reported at each sensitivity value, 
with ribbons as the 95% CI.
**C)** Balanced precision-recall curves for the test sets.
Mean balanced precision is reported at each recall value, 
with ribbons as the 95% CI.
"}
![](figures/ml-performance.png)
:::
